Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-03-28 paper2.qmd
section:4.2; process:write

<!-- 3. The DAGs -->
Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020]. A DAG is a *graph* consisting of nodes connected by edges, where nodes represent random variables. The term *directed* indicates that edges or arrows extend from one node to another, indicating the direction of causal influence. The absence of an edge implies no direct relationship between the nodes. The term *acyclic* means that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs conventionally depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014]. They also display conditioned variables by surrounding them with a square. Here, *conditioning* refers to restricting the statistical analysis to a subset of the population defined by the values of the given variable [@Neal_2020, pp. 32]. Although DAGs often omit exogenous variables for simplicity, the DAGs presented in this section includes exogenous variables to improve clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].

@@

@@

$$
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 \\
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 \\
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
$$

@@

<!-- 1. What are we going to do? -->
The conceptual-population model integrates two *hierarchical structural components* to examine how different teaching methods influence students' writing ability. This approach builds on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24]. Each structural component defines how observed or latent variables affect the primary latent variable of interest [@Everitt_et_al_2010]. Their hierarchical nature enables researchers to test hypotheses while accounting for uncertainties associated with trait estimation (see @sec-theory-issue2), particularly when stimuli exhibit a hierarchical structure (see @sec-theory-issue1b).

<!-- 2 Explaining what we did -->
The top branch of [DAG -@fig-cj09_dag] depicts the first component. This component describes how the students' writing-quality trait $(T_{I})$ and a set of *relevant* text-related variables $(X_{IA})$ (e.g., text length) causally influence the texts' written-quality trait $(T_{IA})$. The DAG also highlights the idiosyncratic errors $(e_{IA})$, which capture variations in the texts' trait that neither the students' trait nor other relevant text-related variables can explain. In this structure, $T_{I}$ and $e_{IA}$ are latent vectors of length $n_{I}$ and $(n_{I} \cdot n_{A})$, respectively. Additionally, $X_{IA}$ is an observed matrix with a dimension of $(n_{I} \cdot n_{A})$ rows and $q_{IA}$ columns (variables).

Additionally, the DAG's top branch illustrates how a set of *relevant* student-related variables $(X_{I})$, such as the teaching method, causally influence the students' writing-quality trait $(T_{I})$. The DAG also includes the idiosyncratic errors $(e_{I})$, which capture variations in the students' traits that the relevant student-related variables cannot explain. Here, $e_{I}$ is a latent vector of length $n_{I}$, and $X_{I}$ can be an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables).

Here, *relevant* refers to variables that satisfy the *backdoor criterion* [@Neal_2020, pp 37], that is, they belong to a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. A *sufficient* set (potentially empty) blocks all non-causal paths between a predictor and an outcome without opening new ones [@Pearl_2009].

@@

Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component. This component describes how the judges' bias $(B_{J})$ and a set of *relevant* judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes) causally influence the biases associated with each text $B_{JK}$. The DAG also highlights the idiosyncratic errors $(e_{JK})$, which capture variations in the texts' bias that neither the judges' bias nor other relevant judgment-related variables can explain. In this description, $B_{J}$ and $e_{JK}$ are latent vectors of length $n_{J}$ and $(n_{J} \cdot n_{K})$, respectively. Additionally, $Z_{JK}$ is an observed matrix with a dimension of $(n_{J} \cdot n_{K})$ rows and $q_{JK}$ columns.

Moreover, the DAG's bottom branch describes how a set of *relevant* judge-related variables $(Z_{J})$, such as judgment expertise, causally influence the judges' bias $(B_{J})$. The DAG also accounts for idiosyncratic errors $(e_{J})$, capturing the variations in the judges' bias that relevant judge-related variables cannot explain. Here, $e_{J}$ is a latent vector of length $n_{J}$, and $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns.

@@

@@

<!-- 1. start by defining the sampling vector and matrices  -->
To facilitate the interpretation of the sample-comparison model, we first define the data sample using several binary vector variables: $S_{I}$, $S_{J}$, $S_{IA}$, $S_{JK}$, and $S$ as follows.

$$
S_{I} = \begin{bmatrix}
i_{(1)} \\
i_{(2)} \\
\vdots \\
i_{(h)} \\
\vdots \\
i_{(i)} \\
\vdots \\
i_{(nI)}
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
j_{(1)} \\
j_{(2)} \\
\vdots \\
j_{(j)} \\
\vdots \\
j_{(nJ)}
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
ia_{(1)} \\
ia_{(2)} \\
\vdots \\
ia_{(ia)} \\
\vdots \\
ia_{(nI \cdot nA)} \\
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
jk_{(1)} \\
jk_{(2)} \\
\vdots \\
jk_{(jk)} \\
\vdots \\
jk_{(nJ \cdot nK)} \\
\end{bmatrix}; \;
S = \begin{bmatrix}
s_{(1)} \\
s_{(2)} \\
\vdots \\
s_{(s)} \\
\vdots \\
s_{(n)} \\
\end{bmatrix}
$$ {#eq-mat21}

Each element of the vectors $S_{I}$ and $S_{J}$ is a binary value that indicates the presence $(1)$ or absence $(0)$ of data in the $I$ and $J$ vectors, as defined in equation ([-@eq-mat11]). Therefore, the vector $S_{I}$ contains $n_{I}$ elements, and the vector $S_{J}$ contains $n_{J}$ elements. For example, we define each element of the vector $S_{I}$ as in equation ([-@eq-mat22]), and we apply a similar definition to $S_{J}$.

@@

Each element of the vectors $S_{IA}$ and $S_{JK}$ is defined like $i_{(i)}$ with vector $I$, but applied to data in the matrices $IA$ and $JK$, as defined in equation ([-@eq-mat11]). Therefore, the vector $S_{IA}$ contains $n_{I} \cdot n_{A}$ elements, and the vector $S_{JK}$ contains $n_{J} \cdot n_{K}$ elements. For example, we define each element of the vector $S_{IA}$ as in equation ([-@eq-mat23]), and we apply a similar definition to $S_{JK}$.

@@

The equation shows that the observed values in the $S_{I}$ and $S_{J}$ vectors determine the binary values in the $S_{IA}$ and $S_{JK}$ vectors, through the $IA$ and $JK$ matrices, respectively. Specifically, researchers cannot observe stimuli and judgments for individuals and judges they did not sample initially.

Finally, the vector $S$ in equation ([-@eq-mat21]) contains $n = \binom{n_{I} \cdot n_{A}}{2} \cdot n_{J} \cdot n_{K}$ elements, with each element defined in equation ([-@eq-mat24]). This equation shows that the observed values in $S_{IA}$ and $S_{JK}$ determine the binary values in $S$ through the $R$ matrix. Specifically, researchers can only observe a subset of judges' comparisons for a subset of stimuli generated by a subset of individuals.

$$
s_{(s)} = \begin{cases} 
1 & \text{if the data for the } i,a \text{ with the } h,b \text{ and the } j,k \text{ elements of } R \text{ are observed} \\
0 & \text{if the data for the } i,a \text{ with the } h,b \text{ and the } j,k \text{ elements of } R \text{ are missing}
\end{cases}
$$ {#eq-mat24}

<!-- 2. What is the effect of the sampling mechanism -->
With the definition of $S$, we can now integrate the sample mechanism into the conceptual-population model. @fig-cj12 illustrates this model. Following the convention of @McElreath_2020 [pp. 499-516] and @Deffner_et_al_2022 [pp. 6], the [DAG -@fig-cj12_dag] depicts the conceptual-population outcome $O^{cp}_{R}$ as an unobserved variable, emphasizing the researchers' inability to access such an outcome due to the sampling mechanism. Additionally, the DAG depicts the conditioned *sample design* variable $S$ causally influencing the sample-comparison outcome $O^{sc}_{R}$. This variable shows that the sample-comparison outcome is a subset drawn from the conceptual-population outcome $O^{cp}_{R}$.

@@

Notice,
<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->

@@

<!-- recording: Sven 24.10.04; time: 00:46:00 - 01:01:00 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:24:50 - 00:30:35 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:49:30 - 00:53:30 -->

@@

@@

::: {#fig-cj15_stat fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{R} & = O^{sc}_{R} \odot C \\
  O^{sc}_{R} & = O^{cp}_{R} \odot S \\
  O^{cp}_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
  D_{R} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
  e_{I} & \overset{iid}{\sim} \text{Normal}(0,1) \\
  e_{J} & \overset{iid}{\sim} \text{Normal}(0,1) \\
  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
\end{aligned}
$$

@@

::: {#fig-cj15_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{R} & := f_{C}(O^{sc}_{R}, C) \\
  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
  O^{cp}_{R} & := f_{O}(D_{R}) \\
  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{I} & \dsep \{ e_{J}, e_{IA}, e_{JK} \} \\
  e_{J} & \dsep \{ e_{IA}, e_{JK} \} \\
  e_{IA} & \dsep e_{JK}
\end{aligned}
$$

Sample-comparison model, including sample and comparison mechanisms and assuming equal discriminal dispersions for the student's traits
:::

::: {#fig-cj16 layout-ncol=2}

::: {#fig-cj16_stat fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{R} & = O^{sc}_{R} \odot C \\
  O^{sc}_{R} & = O^{cp}_{R} \odot S \\
  O^{cp}_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
  D_{R} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
  e_{I} & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
  e_{J} & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK}) \\ \\
  & \textbf{Contraints:} \\
  & \sum_{g=1}^{G_{XI}} s_{XI}/G_{XI} = 1 ; \quad \sum_{g=1}^{G_{ZJ}} s_{ZJ}/G_{ZJ} = 1
\end{aligned}
$$

::: {#fig-cj16_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{R} & := f_{C}(O^{sc}_{R}, C) \\
  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
  O^{cp}_{R} & := f_{O}(D_{R}) \\
  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{I} & := f_{e}(X_{I}) \\
  e_{J} & := f_{e}(Z_{J}) \\
  e_{IA} & \dsep e_{JK}
\end{aligned}
$$

@@

@@

@@

@@

<!-- A cautious reader will notice that the model descriptions of @sec-theory-theoretical_P rely on several relevant variables. In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refers to those that satisfy the *backdoor criterion* [@Neal_2020, pp. 37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].  -->

>>>>

<!-- 3. The DAGs -->
Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020]. A DAG is a *graph* consisting of nodes connected by edges, where nodes represent random variables. The term *directed* indicates that edges or arrows extend from one node to another, indicating the direction of causal influence. The absence of an edge implies no direct relationship between the nodes. The term *acyclic* means that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs conventionally depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014]. Although DAGs often omit exogenous variables for simplicity, the DAGs presented in this section includes exogenous variables to improve clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].

@@

<!-- recording: Sven and Tine 25.03.28; time: 00:10:00 - 00:11:00 -->

@@

$$
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 \\
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 \\
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
$$ {#eq-mat13}

@@

<!-- recording: Sven and Tine 25.03.28; time: 00:12:45 - 00:33:00 -->

<!-- 1. What are we going to do? -->
Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24], the conceptual-population model integrates two *hierarchical structural components* to examine how different teaching methods influence students' writing ability. Each structural component defines how observed or latent variables affect the primary latent variable of interest [@Everitt_et_al_2010]. Their hierarchical nature enables researchers to test hypotheses while accounting for uncertainties associated with trait estimation (see @sec-theory-issue2), particularly when stimuli exhibit a hierarchical structure (see @sec-theory-issue1b).

<!-- 2 Explaining what we did -->
The top branch of [DAG -@fig-cj09_dag] illustrates the first component, where *relevant* [^1] student-related variables $(X_{I})$, such as the teaching method, causally influence the latent variable representing students' writing-quality trait $(T_{I})$. It also shows how $T_{I}$ and *relevant* [^2] text-related variables $(X_{IA})$ (e.g., text length) causally influence the texts' written-quality trait $(T_{IA})$, the primary latent variable of interest. Additionally, the branch accounts for idiosyncratic errors: $e_{I}$ captures unexplained variations in students' traits unexplained by $X_{I}$, while $e_{IA}$ captures variations in texts' traits unexplained by $T_{I}$ or $X_{IA}$. Here, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$. Similarly, $X_{IA}$ is an observed matrix with dimension $(n_{I} \cdot n_{A})$ rows and $q_{IA}$ columns (variables), while $e_{IA}$ and $T_{IA}$ are latent vectors of length $n_{I} \cdot n_{A}$.

[^1]: Here, *relevant* refers to variables that satisfy the *backdoor criterion* [@Neal_2020, pp 37], that is, they belong to a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. A *sufficient* set (potentially empty) blocks all non-causal paths between a predictor and an outcome without opening new ones [@Pearl_2009]. While these topics are beyond the scope of this study, readers seeking a deeper understanding can refer to introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020, as well as introductory books like @Pearl_et_al_2018, @Neal_2020 and @McElreath_2020 useful. For more advanced study, seminal papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, along with books such as @Pearl_2009, @Morgan_et_al_2014 and @Hernan_et_al_2020, are recommended.

[^2]: refer to footnote 1.

@@

Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^3] judge-related variables $(Z_{J})$, such as judgment expertise, causally influence the latent variable representing judges' bias $(B_{J})$. It also shows how $B_{J}$ and *relevant* [^4] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes) causally influence the biases associated with each text $B_{JK}$, the primary latent variable of interest. Additionally, the branch accounts for idiosyncratic errors: $e_{J}$ captures unexplained variations in judges' bias unexplained by $Z_{J}$, while $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$. Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$. Similarly, $Z_{JK}$ is an observed matrix with dimension $(n_{J} \cdot n_{K})$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent vectors of length $n_{J} \cdot n_{K}$.

[^3]: refer to footnote 1.
[^4]: refer to footnote 1.

@@

<!-- recording: Sven and Tine 25.03.28; time: 00:09:00 - 00:10:00 -->
<!-- recording: Sven and Tine 25.03.28; time: 00:11:30 - 00:12:25 -->

@@

<!-- 1. start by defining the sampling vector and matrices  -->
To incorporate the sampling mechanism and facilitate the interpretation of the sample-comparison model, we first define the data sampling process using the binary vector variables $S_{I}$, $S_{J}$, $S_{IA}$, and $S_{JK}$ as follows:

$$
S_{I} = \begin{bmatrix}
i_{(1)} \\
i_{(2)} \\
\vdots \\
i_{(h)} \\
\vdots \\
i_{(i)} \\
\vdots \\
i_{(nI)}
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
j_{(1)} \\
j_{(2)} \\
\vdots \\
j_{(j)} \\
\vdots \\
j_{(nJ)}
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
ia_{(1)} \\
ia_{(2)} \\
\vdots \\
ia_{(ia)} \\
\vdots \\
ia_{(nI \cdot nA)} \\
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
jk_{(1)} \\
jk_{(2)} \\
\vdots \\
jk_{(jk)} \\
\vdots \\
jk_{(nJ \cdot nK)} \\
\end{bmatrix}
$$ {#eq-mat21}

Where each element of the vectors $S_{I}$ and $S_{J}$ is a binary value indicating the presence or absence of data in the $I$ and $J$ vectors (see equation ([-@eq-mat11])). Thus, the vectors $S_{I}$ and $S_{J}$ contains $n_{I}$ and $n_{J}$ elements, respectively. Specifically, we define the vector $S_{I}$ as in equation ([-@eq-mat22]) and apply a similar definition to $S_{J}$.

@@

Similarly, each element of the vectors $S_{IA}$ and $S_{JK}$ is a binary value indicating the presence or absence of data in the matrices $IA$ and $JK$ (see equation ([-@eq-mat11])). Thus, the vectors $S_{IA}$ and $S_{JK}$ contains $n_{I} \cdot n_{A}$ and $n_{J} \cdot n_{K}$ elements, respectively. Specifically, we define the vector $S_{IA}$ as in equation ([-@eq-mat23]) and apply a similar definition to $S_{JK}$.

@@

We can visualize these vectors more clearly using the example in equation ([-@eq-mat13]). Suppose the sample excludes the second student and the third judge. Additionally, assume that the second text from each individual and the third judgment from each judge are also missing. Under these conditions, the vectors have the following structure:

$$
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
0 \\
1 \\
1 \\
0 \\
0 \\
0 \\
0 
\end{bmatrix}
$$ {#eq-mat24}

Equation ([-@eq-mat24]) shows that missing values in the $S_{I}$ and $S_{J}$ vectors directly determine those in the $S_{IA}$ and $S_{JK}$ vectors. In other words, researchers can only observe texts and judgments from students and judges initially included in the sample. Additionally, it shows that the sum of observed elements in $S_{I}$ equals the sampled students $(n^{s}_{I})$ and that a similar sum in vector $S_{J}$ equals the sampled judges $(n^{s}_{J})$. Conversely, the sum of observed elements in $S_{IA}$ represents the total sampled texts across all students $(n^{s}_{I} \cdot n^{s}_{A})$, while a similar sum in vector $S_{JK}$ represents the total sampled judgments across all judges $(n^{s}_{J} \cdot n^{s}_{K})$.

Finally, we define the *sample design* vector $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every component of $S_{JK}$. Essentially, this vector serves as the sampled equivalent of the $R$ matrix defined in equation ([-@eq-mat12]), where the sum of its elements represents the total data sample $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.

$$
S = \begin{bmatrix}
s_{(1)} \\
s_{(2)} \\
\vdots \\
s_{(s)} \\
\vdots \\
s_{(n)} \\
\end{bmatrix}
$$ {#eq-mat25}

<!-- 2. What is the effect of the sampling mechanism -->
With the definition of $S$, we incorporate the sample mechanism into the conceptual-population model. Following the convention of @McElreath_2020 [pp. 499-516] and @Deffner_et_al_2022 [pp. 6], the [DAG -@fig-cj12_dag] represents the conceptual-population outcome $O^{cp}_{R}$ as unobserved, emphasizing that researchers cannot directly access this outcome due to the sampling mechanism.

Additionally, the DAG depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$. A square encloses this vector, indicating that it is a conditioned variable. In this context, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a given variable [@Neal_2020, pp. 32]. Thus, the vector $S$ is responsible for selecting the subset of observation from the conceptual-population outcome $O^{cp}_{R}$, generating the sample-comparison outcome $O^{sc}_{R}$.

@@

@@

<!-- recording: Sven 24.10.04; time: 00:46:00 - 01:01:00 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:24:50 - 00:30:35 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:49:30 - 00:53:30 -->
<!-- recording: Sven and Tine 25.03.28; time: 00:33:55 - 00:37:50 -->

@@

when you talk about the variability, refer that the equal dispersion is just assuming 1 for identification.

@@

::: {#fig-cj15_stat fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{R} & = O^{sc}_{R} \odot C \\
  O^{sc}_{R} & = O^{cp}_{R} \odot S \\
  O^{cp}_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
  D_{R} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
  e_{I} & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
  e_{J} & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK}) \\ \\
  & \textbf{Contraints:} \\
  & \sum_{g=1}^{G_{XI}} s_{XI}/G_{XI} = 1 ; \quad \sum_{g=1}^{G_{ZJ}} s_{ZJ}/G_{ZJ} = 1
\end{aligned}
$$

@@

::: {#fig-cj15_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{R} & := f_{C}(O^{sc}_{R}, C) \\
  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
  O^{cp}_{R} & := f_{O}(D_{R}) \\
  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{I} & := f_{e}(X_{I}) \\
  e_{J} & := f_{e}(Z_{J}) \\
  e_{IA} & \dsep e_{JK}
\end{aligned}
$$

@@

<!-- Use your example, teaching practice and Aptitude-Treatment Interaction (ATI), certain interventions do not have the same effect for all pupils, treatment is a random intercept and slope variable. Relevant variables depends on the context, you really have to think through which variables you have to put in. -->

@@

<!-- remember to mention fidelity of the theory in the model -->

@@

<!-- Causal inference frameworks do not magically guarantee value-free answers, but they force us to be precise about the questions we ask, and to be transparent about the assumptions that we are willing to make [@Rohrer_et_al_2022, pp. 4-5] -->

@@

<!-- *Relevant* refers to variables that satisfy the *backdoor criterion* [@Neal_2020, pp 37], that is, they belong to a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. A *sufficient* set (potentially empty) blocks all non-causal paths between a predictor and an outcome without opening new ones [@Pearl_2009]. Use your example, teaching practice and Aptitude-Treatment Interaction (ATI), certain interventions do not have the same effect for all pupils, treatment is a random intercept and slope variable. Relevant variables depends on the context, you really have to think through which variables you have to put in. -->
